---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions,echo=FALSE}
library(knitr)

opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```


## Loading and preprocessing the data

First of all unzip the data.

```{r}
if(!file.exists("activity.csv")){
  unzip("activity.zip")
}
```

Then load the csv data in a dataframe.

```{r}
rawData <- read.table("activity.csv",sep=",",na.strings="NA",header=TRUE,stringsAsFactors=FALSE)
```

Finally adjust the date to be of type Date, and display a summary to check that everything is fine.

```{r}
rawData1 <- rawData

rawData1[,2] <- as.Date(rawData1[,2],'%Y-%m-%d')

str(rawData1)

```

## What is mean total number of steps taken per day?

```{r}
perDay <- split(rawData1, rawData1$date)

stepsPerDay <- data.frame( day = as.Date(character()), stepsNb = numeric(0))

daysNb <- length(perDay)

for( dayNb in 1:daysNb) {
  stepsPerDay[dayNb,1] <- perDay[[dayNb]]$date[1]
  stepsPerDay[dayNb,2] <- sum( perDay[[dayNb]]$steps, na.rm = TRUE)
}

str(stepsPerDay)

```

```{r fig.align='center', fig.height=4}
hist(stepsPerDay[,2],freq=TRUE,main="Steps per day",xlab="steps",breaks=50)
```

```{r}
stepsMedian = as.integer(median(stepsPerDay[,2]))
stepsMean = round(mean(stepsPerDay[,2]),2)
```

The median of steps per day is **`r stepsMedian`** and the mean of the steps per day is **`r stepsMean`**. 

## What is the average daily activity pattern?

```{r}
#split the data on intervals
rawData2 <- rawData
rawData2[,2] <- as.Date(rawData2[,2],'%Y-%m-%d')
rawData2[,3] <- as.factor(rawData2[,3])
perInterval <- split(rawData2,rawData2$interval)

#initialize the output dataset
averageStepsPerInterval <- data.frame( itv1 = character(0), stepsAverage = numeric(0))
averageStepsPerInterval[,1] <- as.character(averageStepsPerInterval[,1])

#iterate and fullfill the dataset
itvNb <- length(perInterval)
for( i in 1:itvNb) {
  intervalName <- as.character(perInterval[[i]]$interval[1])
  averageStepsPerInterval[i,1] <- intervalName
  averageStepsPerInterval[i,2] <- mean( perInterval[[i]]$steps, na.rm = TRUE)
}

#show a summary to check that everything is ok ...
str(averageStepsPerInterval)

```

```{r fig.height=4}
x <- averageStepsPerInterval[,1]
y <- averageStepsPerInterval[,2]
plot(x,y,type="l",main="Average steps on intervals",ylab="steps",xlab="interval")

maxSteps <- max(averageStepsPerInterval[,2])
maxActivityInterval <- averageStepsPerInterval[averageStepsPerInterval[,2]==maxSteps,1]
```

The interval with the maximum numer of steps is **`r maxActivityInterval`**.

## Imputing missing values

```{r}
#calculate NAs
naCount = length( rawData[is.na(rawData$steps), ]$steps )
```
The number of rows with NA values is **`r naCount`**.

```{r}
#initialize the output dataset
noNAData <- data.frame( steps = numeric(0), date = character(0), interval = numeric(0))
noNAData[,2] <- as.character(noNAData[,2])

rdCt = length(rawData$date)

for( r in 1:rdCt ) {
  noNAData[r,2] = rawData[r,2]
  noNAData[r,3] = rawData[r,3]
  if(is.na(rawData[r,1])) {
    istr = as.character(rawData[r,3])
    noNAData[r,1] = as.integer(averageStepsPerInterval[averageStepsPerInterval[,1]==istr,2])
  } else {
    noNAData[r,1] = rawData[r,1]
  }
}

noNAData[,2] <- as.Date(noNAData[,2],'%Y-%m-%d')

str(noNAData)

```

```{r}
perDay2 <- split(noNAData, noNAData$date)

stepsPerDay2 <- data.frame( day = as.Date(character()), stepsNb = numeric(0))

daysNb <- length(perDay2)

for( dayNb in 1:daysNb) {
  stepsPerDay2[dayNb,1] <- perDay2[[dayNb]]$date[1]
  stepsPerDay2[dayNb,2] <- sum( perDay2[[dayNb]]$steps, na.rm = TRUE)
}

str(stepsPerDay2)

```

```{r fig.align='center', fig.height=4}
hist(stepsPerDay2[,2],freq=TRUE,main="Steps per day",xlab="steps",breaks=50)
```

```{r}
stepsMedian2 = as.integer(median(stepsPerDay2[,2]))
stepsMean2 = format(round(mean(stepsPerDay2[,2]),2),nsmall=2)
```

The median of steps per day for data with no NAs inside is **`r stepsMedian2`** and the mean, for the same dataset is **`r stepsMean2`**. 

The values differ from the case when NAs, there are many days with steps different than 0 than before and thus the mean and median increased and also the histogram looks more natural without that odd peek for days with 0 steps.


## Are there differences in activity patterns between weekdays and weekends?

Create the factor weekday and weekend. Name it 'kind'.

```{r}
sundays = weekdays(noNAData$date) == "Sunday"
saturdays = weekdays(noNAData$date) == "Saturday"
weekendDays = sundays | saturdays
noNAData$kind = "weekday"
noNAData[weekendDays,4] = "weekend"

noNAData$kind = as.factor(noNAData$kind)

str(noNAData)


```


```{r}
library(plyr)

noNAData <- transform(noNAData, interval = as.factor(interval) )

twoDS <- split(noNAData, noNAData$kind)
ds1 <- twoDS[[1]][,c(1,3)]
ds1a <- ddply(ds1,.(interval),summarize,sum=mean(steps))
ds1a$kind = "weekday"

ds2 <- twoDS[[2]][,c(1,3)]
ds2a <- ddply(ds2,.(interval),summarize,sum=mean(steps))
ds2a$kind = "weekend"

finalData = rbind(ds1a,ds2a)
finalData <- transform(finalData, kind = as.factor(kind) )

str(finalData)

library(lattice)
xyplot(sum ~ interval | kind, type="l", ylab="steps average", data = finalData, layout = c(1,2))

```
